Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-07-02 WholePlotTrajectories.tex
maj

\Abstract{
Understanding the ecological rules underlying the maintenance of
tropical forests biodiversity, structure, functioning and dynamics is
urgent to anticipate their fate in the global changing context. The huge
diversity of tropical forests is often assumed to be shaped by constant
regime of disturbance yielding a diversity peak at intermediate
intensity, but for tropical forests this intermediate disturbance
hypothesis (IDH) remains debated, and this controversy also questions
the extent of communities resilience regarding their functional and
taxonomic facets is also to be clarified. To disentangle the ecological
processes driving communities response to disturbance we analysed the
taxonomic and functional diversity trajectories following a logging and
thinning disturbance gradient. Specifically we examined over 30 years
the trajectories of communities taxonomic richness and evennes and
functional composition, diversity and redundancy based on 7 leaf, stem
and life history traits. Trajectories highlighted communities taxonomic
resilience, maintaining their initial differences in composition, and
functional resilience through fast functional trajectory common to all
communities in the functional space. The IDH consistently predicted
communities functional trajectories according to the disturbance
intensity, but poorly represented their taxonomic trajectories, blurried
by the alteration of the functional redundancy and the recruitment of
infrequent, shade tolerant species slowed by competitive exclusion
processes. Although communities functioning had recovered after 30 years
their taxonomic diversity and composition in infrequent species and
their functional redundancy remained altered. these results acknowledged
the need of decades-long recovery cycles to ensure communities complete
resilience, and questioned communities maintenance after repeated
disturbance.
}

@@

In tropical forest, ecological communities are constantly re-shaped by
natural disturbance events that change both the abiotic environment,
through the fluxes of light, heat and water \citep{Goulamoussene2017},
and the biotic interactions like competition among species
\citep{Chesson2000, Herault2018}. One of the cornerstone of tropical
forest ecology is to understand the processes and drivers of ecosystems
response to disturbance \citep{White2001, Chazdon2003a}. For now, this
has been largely studied through forest structural parameters, rapid and
convenient to measure, as aboveground biomass, tree height or stem
density \citep{Piponiot2016, Rutishauser2016}. These structural
parameters were thereafter sucessfully modeled, giving important
insights into the recovery of ecosystems processes and services
\citep{Herault2018}. However the response of forests diversity is still
unclear, albeit it determines the productivity, stability and
functioning of ecosystems \citep[\citet{Liang2016}]{Tilman2014} and
would be most probably impacted by the changes induced by disturbance
\citep{Baraloto2012a}.

In the short-term, moderate disturbance may lead to positive impacts on
communities diversity, which have been formalized by the intermediate
disturbance hypothesis (IDH) stating a maximized species diversity at
intermediate disturbance intensity
\citep{Molino2001, Kariuki2006a, Berry2008a}. Still, validations of the
IDH remain scarce in the long term and mainly rely on the analysis of
species richness that gives limited or misleading information on forests
recovery and functioning \citep{Martin2015, Chaudhary2016}. More
complete analysis would besides encompass communities composition that
is crucial for conservation issues, and abundance distribution to reveal
the ecological rules structuring communities
\citep{Magurran1988, Lavorel2002, Bellwood2006}. Furthermore, a
functional approach accounting for species biological attributes and
role in the ecosystem directly links ecosystems biodiversity,
functioning and environmental constraints
\citep{Violle2007b, Moretti2009, Baraloto2012a, Scheiter2013}. The
functional trait-based approach focused on major traits related to
species ecology and performance was sucessfully adopted
\citep{Diaz2005, Villeger2008a}, for example to reveal in tropical
rainforests the deterministic processes that foster after disturbance
the fast growing species with efficient resources acquisition
\citep{Molino2001, Haddad2008, Ruger2009}. This basic deterministic
process entails a functional shift from a dominance of ``conservative''
slow-growing species dealing with scarce resources to ``acquisitive''
fast-growing species with rapid and efficient use of abundant resources
\citep{TerSteege2001, Reich2014, Herault2011}. This shift is translated
into consistent trajectories of key functional traits related to
resource acquisition (leaf area, density and chlorophyll content, and
stem specific gravity and bark thickness), tree growth and reproduction
life history traits (seed mass and maximum height)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. Besides
communities functional composition and diversity, a last key aspect of
communities response to disturbance is the changes in functional
redundancy that quantifies the amount of shared trait values among
species \citep{Carmona2016}. High functional redundancy, as in the very
diverse tropical forests \citep{Bellwood2006}, mitigates the impacts of
species removal on ecosystem functioning and determines communities
resilience \citep{Trenbath1999, Elmqvist2003, Diaz2005}. High functional
redundancy also increases the functional overlap among species and would
foster neutral stochastic processes \citep{Gravel2006}.

To grasp all facets of communities response to disturbance we examined
here the taxonomic and functional trajectories in terms of richness,
evennes and composition \citep{Lohbeck2015, Guariguata2001}. These
trajectories would highlight the recovery of communities initial
characteristics and the underlying ecological processes. They would
clarify the tenants in the long term of the Intermediate Disturbance
Hypothesis debated in tropical forests, and provide indications for
future adaptive conservation strategies \citep{Adler2007}. We monitored
over 30 years the response of 75 ha of neotropical forest plots set up
on a gradient of disturbance intensity, from 10 to 60\% of ecosystem
biomass removed. We made use of a large functional traits database
browsing major leaf, stem and seed functional traits and species maximum
height to draw the trajectories over time of communities taxonomic and
functional composition,richness and evenness. Specifically, we (i)
questioned the coupling between taxonomic and functional response to
disturbance and identified the underlying assembly processes, (ii)
clarified the validity of the IDH in the long term for tropical forest
and (iii) questioned the completeness of communities recovery regarding
communities functional redundancy.

@@

The plot 7 from treatment 1 displayed a constantly outlying functoinal
evenness and was removed from the graphical representation for better
readability (see appendix for full graphs). In undisturbed plots the
functional evenness remained stable along the 30 years while in
disturbed plots it followed unimodal trajectories with a return towards
initial values starting around 20 years after disturbance.

@@

\begin{quote}
\begin{quote}
Si on met que richesse et evenness, pour functional c'est de l'evenness
(Rao = order2)
\end{quote}
\end{quote}

@@

The taxonomic differences among communities, marked before disturbance
by the distinct starting points of the NMDS, were maintained throughout
communities recovery trajectories. More than commonly thought
post-disturbance trajectories depended on communities initial
composition, that partly determined the pool of rectuitable species and
that constrained the converge towards the recovery of the initial
composition
\citep{Hubbell1999, Molino2001, Anderson2007, Baraloto2012a}. The
resilience of communities taxonomy suggested the existence of multiple
stable equilibria, here the pre-disturbance communities, as assumed for
highly diverse and productive ecosystems \citep{Chase2003}.

@@

The humped-shaped trajectories of communities along time demonstrated
the enhanced functional evennes and changes in functional composition
following disturbance before a return towards initial, lower values. In
accordance with the IDH, disturbance entailed significant temporal
variability with the emergence of ecological processes different from
those rueling undisturbed forest. The increase in functional evenness
went with significant functional shifts towards resource-acquisitive
strategies (sharp increase in the SLA, leaf thickness and bark thickness
and decrease in wood density, leaf toughness and maximum height)
\citep{Westoby1998, Wright2004, Reich2014}. Because the pre-disturbance
survivor trees mirror the initial communities \citep{Herault2018}, the
functional changes relied upon recruited trees and the enhanced growth
and survival of previously infrequent species and functional types.
Pioneers outcompeted other species in the post-disturbance environment
of non limiting resources and alleviated competition. Before disturbance
and over communities recovery time though, these species would be
excluded by long-lived, more resistant and shade tolerant species.

@@

The recovery of the most infrequent species followed the lottery
recruitment rules \citep{Busing2002} and was hampered by the increasing
competition among species whih entailed long term recovery cycles
\citep{Trenbath1999, Elmqvist2003, Diaz2005}.

The rapid decrease of the functional redundancy within the initial
functional space, correlated with the disturbance intensity, illustrated
the space and resources made available after disturbance. Following the
lottery recruitment, communities taxonomic richness increased firsthand
along the enhanced recruitment rate but decreased thereafter, following
the emergence of exclusive competition. After 30 years the recovery of
the initial redundancy structure was underway but remained unachieved.
This alteration of the functional redundancy meant a lower resilience of
pre-disturbance communities and higher chances to see the persistence of
disturbance-specific communities, with less species and more pioneers
\citep{Haddad2008, Burslem2000, Martin2013}. Besides the hampered
recovery of rare species increased the risks to loose cornerstone
specie, with unexpected ecological consequences
\citep{Jones1994, Chazdon2003a, Diaz2005, Gardner2007}. Apart from the
functional characteristics considered here, infrequent species mighthave
unique functions i nthe ecosystem or be a key for some fauna.

@@

Our study revealed communities cyclic recovery after disturbance
entailing the resilience of their functioning and taxonomic composition
with the maintenance of initial differences among communities.
Communities functional evenness was enhanced for 20 years after
disturbance through the enrichment of the communities with pioneers and
light-demanding species,in accordance with the IDH. The IDH, though,
poorly predicted the disturbance impact on communities taxonomic
richness and evenness that were blurried by the emergence of competitive
exclusion along time. The resilience of tropical forests proved
consistent although spread over several decades. Still, the disturbance
impact on communities redundancy cautioned against the risks of
infrequent species loss and the persistence of disturbance-specific
communities \citep{Gourlet-Fleury2005}. As the trajectories highlighted
the recruitment processes proved central for communities response to
disturbance and closer focus on demographical drivers of communities
response would clarify the fate of the future forests.

>>>>

\Abstract{
Understanding the ecological rules underlying the maintenance of
tropical forests biodiversity, structure, functioning and dynamics is
urgent to anticipate their fate in the global changing context. The huge
diversity of tropical forests is often assumed to be shaped by constant
regime of disturbance yielding a diversity peak at intermediate
intensity, but for tropical forests this intermediate disturbance
hypothesis (IDH) remains debated, and this controversy also questions
the extent of communities resilience regarding their functional and
taxonomic facets is also to be clarified. To disentangle the ecological
processes driving communities response to disturbance we analysed the
taxonomic and functional diversity trajectories following a logging and
thinning disturbance gradient. Specifically we examined over 30 years
the trajectories of communities taxonomic richness and evennes and
functional composition, diversity and redundancy based on 7 leaf, stem
and life history traits. Trajectories highlighted the cyclic recovery of
communities taxonomic composition, maintaining the pre-disturbance
differences, as well as functional composition and diversity. The IDH
predicted communities functional response to disturbance that resulted
from competition processes determined by the disturbance intensity. The
IDH however poorly represented the taxonomic response that relied upon
lottery recruitment and recruitment limitation which hampered the
recovery of infrequent, shade tolerant species. After disturbance
communities functioning followed a cyclic recovery and the initial
composition differences among plots were maintained. Although consistent
the recovery of communities functioning and taxonomic composition and
diversity remained unachieved after 30 years which acknowledged the need
of decades-long recovery cycles to ensure a complete recovery, and
questioned communities sustainability after repeated disturbance.
}

@@

In tropical forests, ecological communities are constantly re-shaped by
natural disturbance events that change both the abiotic environment,
through the fluxes of light, heat and water \citep{Goulamoussene2017},
and the biotic interactions like competition among species
\citep{Chesson2000, Herault2018}. One of the cornerstone of tropical
forest ecology is to understand the processes and drivers of ecosystems
response to disturbance \citep{White2001, Chazdon2003a}. For now, this
has been largely studied through forest structural parameters, rapid and
convenient to measure, as aboveground biomass, tree height or stem
density \citep{Piponiot2016, Rutishauser2016}. These structural
parameters were thereafter sucessfully modeled, giving important
insights into the recovery of ecosystems processes and services
\citep{Herault2018}. However the response of forests diversity is still
unclear, albeit it determines the productivity, stability and
functioning of ecosystems \citep[\citet{Liang2016}]{Tilman2014} and
would be most probably impacted by the changes induced by disturbance
\citep{Baraloto2012a}. In the short-term, moderate disturbance may lead
to positive impacts on communities diversity, which have been formalized
by the intermediate disturbance hypothesis (IDH) stating a maximized
species diversity when the disturbance regime is moderate but
significant \citep{Molino2001, Kariuki2006a, Berry2008a}. Along time the
IDH would translate into a variety of post-disturbance trajectories
shaped by the different ecological processes at stake.

Validations of the IDH though remain scarce in the long term and mainly
rely on the analysis of species richness. Species richness though gives
limited or misleading information on forests recovery and functioning
\citep{Martin2015, Chaudhary2016}: more complete analysis would
encompass communities composition that is crucial for conservation
issues, and abundance distribution that reveals the ecological processes
involved \citep{Magurran1988, Lavorel2002, Bellwood2006}. Furthermore, a
functional approach accounting for species biological attributes and
assessing their role in the ecosystem would directly link communities
diversity to the ecosystem functioning and to its environmental
constraints
\citep{Violle2007b, Moretti2009, Baraloto2012a, Scheiter2013}. In that
respect, the functional trait-based approach that focused on major
traits related to species ecology and performance was sucessfully
adopted \citep{Diaz2005, Villeger2008a}. For example the functoinal
approach revealed in tropical rainforests the deterministic processes
fostering fast growing species with efficient resources acquisition
after disturbance \citep{Molino2001, Haddad2008, Ruger2009}. Disturbance
therefore entails a functional shift from a dominance of
``conservative'' slow-growing species dealing with scarce resources to
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{TerSteege2001, Reich2014, Herault2011}. This
shift is translated into consistent trajectories of key functional
traits related to resource acquisition (leaf area, density and
chlorophyll content, and stem specific gravity and bark thickness), tree
growth and reproduction life history traits (seed mass and maximum
height) \citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}.
Eventually a complete overview of communities response to disturbance
would encompass the changes in functional redundancy, that quantifies
the amount of shared trait values among species \citep{Carmona2016}.
High functional redundancy like in the very diverse tropical forests
\citep{Bellwood2006} mitigates the impacts of species removal on
ecosystem functioning and determines the resilience of communities
\citep{Trenbath1999, Elmqvist2003, Diaz2005}. High functional redundancy
also increases the functional overlap among species and would foster
neutral stochastic processes \citep{Gravel2006}.

To grasp all facets of communities response to disturbance we examined
here the taxonomic and functional trajectories in terms of richness,
evenness, composition and redundancy
\citep{Lohbeck2015, Guariguata2001}. These trajectories would highlight
the recovery of communities initial characteristics and the underlying
ecological processes, and provide indications for future adaptive
conservation strategies \citep{Adler2007}. They would clarify the
tenants of the Intermediate Disturbance Hypothesis in the long term,
debated in tropical forests, and explicit the variety of
post-disturbance trajectories defined by the disturbance intensity. We
monitored over 30 years the response of 75 ha of neotropical forest
plots set up on a gradient of disturbance intensity, from 10 to 60\% of
ecosystem biomass removed. We made use of a large functional traits
database browsing major leaf, stem and seed functional traits and
species maximum height to draw the trajectories over time of communities
taxonomic and functional composition, richness, evenness and redundancy.
Specifically, we (i) questioned the recovery of communities taxonomic
and functional characteristics and identify the underlying assembly
processes, (ii) clarified the validity of the IDH in the long term for
tropical forest and elucidated its declination into different
trajectories, and (iii) questioned the length of complete communities
recovery.

@@

The plot 7 from treatment 1 displayed a constantly outlying functional
evenness and was removed from the graphical representation for better
readability (see appendix for full graphs). In undisturbed plots the
functional evenness remained stable along the 30 years while in
disturbed plots it followed unimodal trajectories with a return towards
initial values starting around 20 years after disturbance.

@@

@@

The taxonomic differences among communities, marked before disturbance
by the distinct starting points of the NMDS, were maintained throughout
communities recovery trajectories. More than commonly thought
post-disturbance trajectories depended on communities initial
composition, that partly determined the pool of recruitable species and
that constrained the converge towards the recovery of the initial
composition
\citep{Hubbell1999, Molino2001, Anderson2007, Baraloto2012a}. The
resilience of communities taxonomy suggested the existence of multiple
stable equilibria, here the pre-disturbance communities, as assumed for
highly diverse and productive ecosystems \citep{Chase2003}.

@@

The humped-shaped trajectories of communities along time illustrated the
processes enhancing the functional evennes and the changes in functional
composition following disturbance before communities return towards
initial composition and lower evenness. Disturbance entailed significant
variability in the communities through the emergence of ecological
processes different from thoses involvemnt before disturbance, in
accordance with the IDH. The increase in functional evenness went with
significant functional shifts towards resource-acquisitive strategies
(sharp increase in the SLA, leaf thickness and bark thickness and
decrease in wood density, leaf toughness and maximum height)
\citep{Westoby1998, Wright2004, Reich2014}. Because the pre-disturbance
survivor trees mirror the initial communities \citep{Herault2018}, the
functional changes relied upon recruited trees and the enhanced growth
and survival of previously infrequent species and functional types.
Disturbance vacated environmental niches of high light, space and
nutrient availability, filled by competitive pioneers that became
dominant and determined the functional characteristics of the community
\citep{Grime1998}. Before disturbance and over communities recovery time
though, these species would be excluded by long-lived, more resistant
and shade tolerant species.

@@

The recovery of the most infrequent species followed the lottery
recruitment rules \citep{Busing2002} and was hampered by the increasing
competition among species, implying long term recovery
\citep{Trenbath1999, Elmqvist2003, Diaz2005}.

Following disturbance the space and resources made available
corresponded to a rapid decrease of the functional redundancy within the
initial functional space. The high light, space and nutrient niches were
rapidly filled by species, mainly the most dominant and frequent ones.
Thereafter competitive exclusion emerged following the filling of the
niches and limited the recruitment of infrequent species
\citep{Busing2002}. The recovery of communities taxonomic richness and
evenness relied upon the recovery of the initial functional redudancy.
Although underway 30 years after disturbance the recovery remained
unachieved and after 30 years the recovery it was underway but remained
unachieved. This alteration of the functional redundancy meant a lower
resilience of pre-disturbance communities and higher chances to see the
persistence of disturbance-specific communities, with less species and
more pioneers \citep{Haddad2008, Burslem2000, Martin2013}. Besides the
slowed recovery of rare species increased the risks to loose cornerstone
specie, with unexpected ecological consequences
\citep{Jones1994, Chazdon2003a, Diaz2005, Gardner2007}. Apart from the
functional characteristics considered here, infrequent species mighthave
unique functions i nthe ecosystem or be a key for some fauna.

@@

Our study revealed communities cyclic recovery after disturbance
allowing the resilience of their functioning and taxonomic composition
with the maintenance of initial differences among communities.
Communities functional evenness was enhanced for 20 years after
disturbance through the enrichment of the communities with pioneers and
light-demanding species,in accordance with the IDH. The IDH, though,
poorly predicted the disturbance impact on communities taxonomic
richness and evenness that were blurried by the emergence of competitive
exclusion along time. The resilience of tropical forests proved
consistent although spread over several decades. Still, the disturbance
impact on communities redundancy cautioned against the risks of
infrequent species loss and the persistence of disturbance-specific
communities \citep{Gourlet-Fleury2005}. As the trajectories highlighted
the recruitment processes proved central for communities response to
disturbance and closer focus on demographical drivers of communities
response would clarify the fate of the future forests.
